First Order Alternation

### 2018-10-05 quantifiers/draft.tex
ok

\begin{definition}\label{def:generalized-lyndon-interpolant}
  Given a non-empty sequence of input events $\alpha = a_1 \ldots a_n
  \in \Sigma^*$, a \emph{generalized Lyndon interpolant (GLI)} is a
  sequence $(I_0,\ldots,I_n)$ of formulae such that, for all $k \in
  [n-1]$: \begin{compactenum}
    \item $\pset{-}{I_k} = \emptyset$,
      %
    \item $\stamp{\iota}{0} \models I_0$ and \(I_k \wedge
      \Big(\bigwedge_{q(\vec{y}) \arrow{a_i(X)}{} \psi \in \Delta}
      \forall y_1 \ldots \forall y_{\#(q)} ~.~ \stamp{q}{k}(\vec{y})
      \rightarrow \stamp{\psi}{k+1}\Big) \models I_{k+1}\),
      %
    \item $I_n \wedge \bigwedge_{q \in Q \setminus F} \forall y_1
      \ldots \forall y_{\#(q)} ~.~ q(\vec{y})$ is unsatisfiable.
  \end{compactenum}
  Moreover, the GLI is \emph{local} if and only if $\voc{I_k}
  \subseteq \stamp{Q}{k}$, for all $k \in [n]$.
\end{definition}
The following proposition states the existence of local GLI for the
theories in which Lyndon's Interpolation Theorem holds.

@@

\begin{example}[Contd. from Examples \ref{ex:quant-elim} and \ref{ex:pred-elim}]\label{ex:gli}
The formula $\substform{\alpha}$ (Example \ref{ex:pred-elim}) is
unsatisfiable and let $\tau_2 = z_1$ be the witness term for the
universally quantified variable $z_2$. Replacing $z_2$ with $\tau_2$
in the matrix of $\quantform{\alpha}$ (Example \ref{ex:quant-elim})
yields the unsatisfiable conjunction: 
\[\begin{array}{l}
z_1 \geq 0 \wedge \stamp{q}{0}(z_1) ~\wedge \\
\stamp{q}{0}(z_1) \rightarrow \stamp{x}{1} \geq 0 \wedge (z_2 \geq z_1 \rightarrow \stamp{q}{1}(\stamp{x}{1}+z_2)) ~\wedge \\
\stamp{q}{1}(\stamp{x}{1}+z_2) \rightarrow \stamp{x}{1}+z_2 < 0 \wedge \stamp{q_f}{2}(\stamp{x}{2}+\stamp{x}{1}+z_2)
\end{array}\]
A non-local GLI for the above is $(\stamp{q}{0}(z_1) \wedge
z_1\geq 0,~ \stamp{x}{1} \geq 0 \wedge \stamp{q}{1}(\stamp{x}{1}+z_1)
\wedge z_1\geq0,~\bot)$. \hfill$\blacksquare$
\end{example}

@@

In conclusion, under two assumptions about the first-order theory of
the data domain, namely \begin{inparaenum}[(i)]
\item witness-producing quantifier elimination, and
\item Lyndon interpolation for the quantifier-free fragment with
  uninterpreted functions,
\end{inparaenum}
we developped a rather generic method that produces generalized Lyndon
interpolants for unfeasible input event sequences. Moreover, each
formula $I_k$ in the interpolant refers only to the current predicate
symbols $\stamp{Q}{I_k}$, the current and past input variables
$\stamp{X}{\leq k}$ and the existentially quantified transition
variables introduced at the previous steps $\set{x_j \mid j <
  \max(\xi^{-1}(k)), Q_j = \exists}$. The remaining question is how to
use such non-local interpolants for the coverage between the nodes of
the unfolding of an automaton (Definition \ref{def:coverage}).

\subsection{Coverage with Non-local Interpolants}

>>>>

\begin{definition}\label{def:generalized-lyndon-interpolant}
  Given a non-empty sequence of input events $\alpha = a_1 \ldots a_n
  \in \Sigma^*$, a \emph{generalized Lyndon interpolant (GLI)} is a
  sequence $(I_0,\ldots,I_n)$ of formulae such that, for all $k \in
  [n-1]$: \begin{compactenum}
    \item\label{it1:generalized-lyndon-interpolant} $\pset{-}{I_k} =
      \emptyset$,
      %
    \item\label{it2:generalized-lyndon-interpolant} $\stamp{\iota}{0}
      \models I_0$ and \(I_k \wedge \Big(\bigwedge_{q(\vec{y})
        \arrow{a_i(X)}{} \psi \in \Delta} \forall y_1 \ldots \forall
      y_{\#(q)} ~.~ \stamp{q}{k}(\vec{y}) \rightarrow
      \stamp{\psi}{k+1}\Big) \models I_{k+1}\),
      %
    \item\label{it3:generalized-lyndon-interpolant} $I_n \wedge
      \bigwedge_{q \in Q \setminus F} \forall y_1 \ldots \forall
      y_{\#(q)} ~.~ q(\vec{y})$ is unsatisfiable.
  \end{compactenum}
  Moreover, the GLI is \emph{local} if and only if $\voc{I_k}
  \subseteq \stamp{Q}{k}$, for all $k \in [n]$.
\end{definition}
The following proposition states the existence of local GLI for the
theories in which Lyndon's Interpolation Theorem holds.

@@

\begin{example}[Contd. from Examples \ref{ex:quant-elim} and \ref{ex:pred-elim}]\label{ex:gli}
The formula $\substform{\alpha}$ (Example \ref{ex:pred-elim}) is
unsatisfiable and let $\tau_2 = z_1$ be the witness term for the
universally quantified variable $z_2$. Replacing $z_2$ with $\tau_2$
in the matrix of $\quantform{\alpha}$ (Example \ref{ex:quant-elim})
yields the unsatisfiable conjunction: 
\[\begin{array}{c}
z_1 \geq 0 \wedge \stamp{q}{0}(z_1)  \\ \wedge \\
\stamp{q}{0}(z_1) \rightarrow \stamp{x}{1} \geq 0 \wedge (z_1 \geq z_1 \rightarrow \stamp{q}{1}(\stamp{x}{1}+z_1)) \\ \wedge \\
\stamp{q}{1}(\stamp{x}{1}+z_1) \rightarrow \stamp{x}{1}+z_1 < 0 \wedge \stamp{q_f}{2}(\stamp{x}{2}+\stamp{x}{1}+z_1)
\end{array}\]
A non-local GLI for the above is $(\stamp{q}{0}(z_1) \wedge
z_1\geq 0,~ \stamp{x}{1} \geq 0 \wedge \stamp{q}{1}(\stamp{x}{1}+z_1)
\wedge z_1\geq0,~\bot)$. \hfill$\blacksquare$
\end{example}

@@

In conclusion, under two assumptions about the first-order theory of
the data domain, namely\begin{inparaenum}[(i)]
\item witness-producing quantifier elimination, and
\item Lyndon interpolation for the quantifier-free fragment with
  uninterpreted functions,
\end{inparaenum}
we developped a rather generic method that produces generalized Lyndon
interpolants for unfeasible input event sequences. Moreover, each
formula $I_k$ in the interpolant refers only to the current predicate
symbols $\stamp{Q}{I_k}$, the current and past input variables
$\stamp{X}{\leq k}$ and the existentially quantified transition
variables introduced at the previous steps $\set{x_j \mid j <
  \max(\xi^{-1}(k)), Q_j = \exists}$. The remaining question is how to
use such non-local interpolants to label the unfolding of an automaton
(Definition \ref{def:unfolding}) and to compute the coverage between
nodes of the unfolding (Definition \ref{def:coverage}).

\subsection{Unfolding with Non-local Interpolants}

As required by Definition \ref{def:unfolding}, the unfolding $U$ of an
automaton $\A = \tuple{\Sigma,X,Q,\iota,F,\Delta}$ is labeled by
formulae $U(\alpha) \in \posforms(Q,\emptyset)$, with no free symbols,
other than predicate symbols, such that the labeling is compatible
with the transition relation of the automaton (point
\ref{it3:unfolding} of Definition \ref{def:unfolding}).

Suppose that a given input event sequence $\alpha \in \dom(U)$ is
labeled with a sequence of formulae $U(\alpha_0) = \iota, U(\alpha_1),
\ldots U(\alpha_n) = U(\alpha)$ in a way that fulfills the requirement
from Definition \ref{def:unfolding} (\ref{it3:unfolding}) and,
moreover, it has a non-local GLI $(I_0, \ldots, I_n)$, such that
$\voc{I_k} \subseteq \stamp{Q}{k} \cup \stamp{X}{\leq k} \cup
\set{x_{i_1},\ldots,x_{i_k}}$, where $x_{i_1},\ldots,x_{i_n}$ are the
renamed variables occurring in the rules of $\Delta$ under the scope
of existential transition quantifiers (formally these are the
existentially quantified variables from the prenex normal form of
$\quantform{\alpha_k}$). The interpolant is used in refining the
labeling of the sequence as follows, for all $k \in [n]$: \begin{compactitem}
\item let $J_k \in \posforms(Q,\emptyset)$ be the formula obtained
  from $I_k$ by\begin{inparaenum}[(i)]
    \item replacing each time-stamped predicate symbol $\stamp{q}{k}$ by $q$, and
      %
    \item existentially quantifying each variable $y \in
      \stamp{X}{\leq k} \cup \set{x_{i_1},\ldots,x_{i_k}}$.
  \end{inparaenum}
%
\item define the new labeling as $U'(\alpha_k) = U(\alpha_k) \wedge J_k$. 
\end{compactitem}
To see that the new set of formulae $U'(\alpha_0), \ldots,
U'(\alpha_n)$ complies with Definition \ref{def:unfolding}, it is
enough to notice the following facts: \begin{compactitem}
\item $U'(\alpha_0) \equiv \iota$, because\footnote{By point
  \ref{it2:generalized-lyndon-interpolant} of Definition
  \ref{def:generalized-lyndon-interpolant}.} $\stamp{\iota}{0} \models
  I_0$, thus $\iota \models J_0$ and $U'(\alpha_0) = U(\alpha_0)
  \wedge J_0 \equiv \iota \wedge J_0$.
%
\item 
\end{compactitem}
